### 2018-06-11 paper/3-abstractions.tex
Work on S3

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10     B1: A1 + A2
A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _     _    = Nothing
\end{minted}
\vspace{1mm}
We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
\emph{callback} \hs{fetch} and a key. It pattern-matches on the key to see if it
has a task description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
it computes the value of the formula, using \hs{fetch} to find the value of any
keys on which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures are given in Fig.~\ref{fig-types}. We will
explain why in subsection~\S\ref{sec-general-build}.

For now, we content ourselves with observing that a task description,
of type \hs{Task}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
compilers, calculation chains, file systems, caches, and all other
complexities of real build systems.  It just computes a single output, in
a side-effect-free way, using a callback (\hs{fetch}) to find the values
of its dependencies.

@@

Next comes our second main abstraction -- a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up to date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: Eq k => Build Monad () k v
busy task key store = execState (fetch key) store
  where
    fetch :: k -> State (Store () k v) v
    fetch k = case task fetch k of
        Nothing  -> gets (getValue k)
        Just act -> do v <- act; modify (putValue k v); return v
\end{minted}
\vspace{1mm}

\noindent
The \hs{busy} build system defines the callback \hs{fetch} so that, when given a
key, it brings the key up to date in the store, and returns its value.
The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
Fig.~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
To bring a key up to date, \hs{fetch} asks the task description \hs{task} how
to compute the value of \hs{k}. If \hs{task} returns \hs{Nothing} the key is an
input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
runs the action \hs{act} returned by the \hs{task} to produce a resulting
value~\hs{v}, records the new key/value mapping in the store, and returns \hs{v}.
Notice that \hs{fetch} passes itself to \hs{task} as an argument, so that the
latter can use \hs{fetch} to recursively find the values of \hs{k}'s dependencies.

@@

\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store  = initialise () (\key -> if key == "A1" then 10 else 20)
@\ghci@ result = busy sprsh1 "B2" store
@\ghci@ getValue result "B1"
30
@\ghci@ getValue result "B2"
60
\end{minted}

@@

The previous example shows why the \hs{Task} abstraction is polymorphic in \hs{f}, recall the definition:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
The \hs{busy} build system instantiates \hs{f} to
\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)},
so that \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} can side-effect the
\hs{Store}, thereby allowing successive calls to \hs{fetch} to communicate with
one another.

@@

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emerged, we find the that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula may depend
  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
  value of another cell \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
  even use the application operator \hs{<*>}, which limits dependencies to a
  single linear chain. It is interesting to note that, when run on a
  \hs{Task}~\hs{Functor}, the \hs{busy} build system will build each key at most
  once, thus partially fulfilling the minimality requirement~\ref{def-minimal}.
  Alas, it still has no mechanism to decide which input keys changed since the
  previous build.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\end{itemize}

Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
argument, an application of \hs{busy} to a \hs{Task}~\hs{Functor} or
a \hs{Task}~\hs{Applicative} will typecheck and run just fine. It feels a bit like
sub-typing, but is actually just ordinary higher-rank polymorphism at
work~\cite{jones2007practical}.

@@

As explained in~\S\ref{sec-background-excel}, some task descriptions have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such task descriptions correspond to the type
\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Consider this spreadsheet example:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh2 :: Task Monad String Integer
sprsh2 fetch "B1" = Just $ do c1 <- fetch "C1"
                              if c1 == 1 then fetch "B2" else fetch "A2"
sprsh2 fetch "B2" = Just $ do c1 <- fetch "C1"
                              if c1 == 1 then fetch "A1" else fetch "B1"
sprsh2 _     _    = Nothing
\end{minted}
\vspace{1mm}

@@

To express this formally we need an auxiliary function \hs{compute}, that
computes the value of a key in a given store \emph{without attempting to update
any dependencies}:
\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> k -> Maybe v
compute task store = fmap runIdentity@\,@.@\,\blk{task}@ (\k -> Identity (getValue k store))
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)}
has the type \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a
\hs{task}. The result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now
need to get rid of the \hs{Identity} wrapper by applying \hs{runIdentity} to the
contents of \hs{Maybe}.

\vspace{-1mm}
\definition[Correctness]{Suppose \hs{build} is a build system, \hs{task} is a
build task description, \hs{key} is a target key, \hs{store} is an initial
store, and \hs{result} is the store produced by running the build system with
parameters \hs{task}, \hs{key} and \hs{store}. Or, using the precise language of
our abstractions:

\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
task          :: Task c k v
key           :: k
store, @@result :: Store i k v
result = @@build @@task @@key @@store
\end{minted}

@@

    \item The \hs{result} is \emph{consistent} with the \hs{task}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{@\std{fromJust}@}~\hs{(}\hs{compute}~\hs{task}~\hs{result}~\hs{k)}
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{2mm}

It is hard to satisfy the above definition of correctness given a task
description with cycles. All build systems discussed in this paper are correct
only under the assumption that the given task description is acyclic. This
includes the \hs{busy} build system introduced earlier: it will loop
indefinitely given a cyclic \hs{task}. Some build systems provide a limited
support for cyclic tasks, see~\S\ref{sec-iterative-compute}.

>>>>

% Old type:
% type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}

\noindent
Here \hs{Task} describes a single build task, while \hs{Tasks} associates a
task description to every non-input key; input keys are associated with
\hs{Nothing}. The highly-abstracted type\footnote{Readers familiar with
\emph{lenses} or \emph{profunctor optics} might recognise a familiar pattern.
We discuss this in~\S\ref{sec-related-optics}.} \hs{Task} is best introduced by
an example. Consider this \Excel spreadsheet:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10     B1: A1 + A2
A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}

\noindent
Here cell \cmd{A1} contains the value \cmd{10}, cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Tasks Applicative String Integer
sprsh1 "B1" = Just $ Task $ \fetch -> ((+)  <$> fetch "A1" <*> fetch "A2")
sprsh1 "B2" = Just $ Task $ \fetch -> ((*2) <$> fetch "B1")
sprsh1 _    = Nothing
\end{minted}
\vspace{1mm}

\noindent
We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.) The
task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values. It pattern-matches on the key to see if it has a task
description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating that the key is an input. If there is a formula in the
cell, it returns the \hs{Task} to compute the value of the formula. Every
\hs{Task} is given a \emph{callback} \hs{fetch} to find the value of any keys on
which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit
mysterious because it takes place in an \hs{Applicative}
computation~\cite{mcbride2008applicative} -- the relevant type signatures are
given in Fig.~\ref{fig-types}. We will explain why in~\S\ref{sec-general-build}.

For now, we content ourselves with observing that a task description, of type
\hs{Tasks}~\hs{c}~\hs{k}~\hs{v}, is completely isolated from the world of
compilers, calculation chains, file systems, caches, and all other
complexities of real build systems.  It just computes a single output, in
a side-effect-free way, using a callback (\hs{fetch}) to find the values
of its dependencies.

@@

Next comes our second main abstraction -- a build system:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}

\noindent
The signature is very straightforward. Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up to date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: Eq k => Build Applicative () k v
busy tasks key store = execState (fetch key) store
  where
    fetch :: k -> State (Store () k v) v
    fetch k = case tasks k of
        Nothing   -> gets (getValue k)
        Just task -> do v <- run task fetch; modify (putValue k v); return v
\end{minted}
\vspace{1mm}

\noindent
The \hs{busy} build system defines the callback \hs{fetch} so that, when given a
key, it brings the key up to date in the store, and returns its value.
The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
Fig.~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
To bring a key up to date, \hs{fetch} asks the task description \hs{tasks} how
to compute the value of \hs{k}. If \hs{tasks} returns \hs{Nothing} the key is an
input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
runs the obtained \hs{task} to produce a resulting value~\hs{v}, records the new
key/value mapping in the store, and returns \hs{v}. Notice that \hs{fetch}
passes itself to \hs{task} as an argument, so that the latter can use \hs{fetch}
to recursively find the values of \hs{k}'s dependencies.

@@

\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store  = initialise () (\key -> if key == "A1" then 10 else 20)
@\ghci@ result = busy sprsh1 "B2" store
@\ghci@ getValue "B1" result
30
@\ghci@ getValue "B2" result
60
\end{minted}

@@

The previous example shows why the \hs{Task} abstraction is polymorphic in \hs{f}, recall the definition:

\begin{minted}[xleftmargin=10pt]{haskell}
newtype Task c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
\end{minted}

\noindent
The \hs{busy} build system instantiates \hs{f} to
\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)},
so that \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} can side-effect the
\hs{Store}, thereby allowing successive calls to \hs{fetch} to communicate with
one another.

@@

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emerged, we find that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula may depend
  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
  value of another cell \cmd{D}.
%TODO: Add a reference for Docker?
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
  even use the application operator \hs{<*>}, which limits dependencies to a
  single linear chain, as e.g. in the Docker build system. It is interesting to
  note that, when run on such a task description, the \hs{busy} build system
  will build each key at most once, thus partially fulfilling the minimality
  requirement~\ref{def-minimal}. Alas, it still has no mechanism to decide which
  input keys changed since the previous build.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

% TODO: Shall we describe the lifTask and liftTasks combinators? They aren't pretty.
%
%   liftTask :: Task Applicative k v -> Task Monad k v
%   liftTask (Task task) = Task task
%
%   liftTasks :: Tasks Applicative k v -> Tasks Monad k v
%   liftTasks = fmap (fmap liftTask)
%
% Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
% argument, an application of \hs{busy} to a \hs{Task}~\hs{Functor} or
% a \hs{Task}~\hs{Applicative} will typecheck and run just fine. It feels a bit like
% sub-typing, but is actually just ordinary higher-rank polymorphism at
% work~\cite{jones2007practical}.

@@

As explained in~\S\ref{sec-background-excel}, some task descriptions have
dynamic dependencies, which are determined by values of intermediate
computations. In our framework, such task descriptions correspond to the type
\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{v}. Consider this spreadsheet example:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh2 :: Tasks Monad String Integer
sprsh2 "B1" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "B2" else fetch "A2"
sprsh2 "B2" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    if c1 == 1 then fetch "A1" else fetch "B1"
sprsh2 _ = Nothing
\end{minted}
\vspace{1mm}

@@

To express this formally we need an auxiliary function \hs{compute}, that
computes the value of a key in a given store \emph{without attempting to update
any dependencies}:

\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.

\vspace{-1mm}
\definition[Correctness]{Suppose \hs{build} is a build system, \hs{tasks} is a
build task description, \hs{key} is a target key, \hs{store} is an initial
store, and \hs{result} is the store produced by running the build system with
parameters \hs{tasks}, \hs{key} and \hs{store}. Or, using the precise language of
our abstractions:

\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
tasks         :: Tasks c k v
key           :: k
store, @@result :: Store i k v
result = @@build @@tasks @@key @@store
\end{minted}

@@

    \item The \hs{result} is \emph{consistent} with the \hs{tasks}, i.e.
    for all non-input keys \hs{k}~$\in$~$O$, the result of recomputing the
    corresponding \hs{task} matches the value stored in the \hs{result}:
    \[
    \hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result}.
    \]
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{tasks}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{2mm}

It is hard to satisfy the above definition of correctness given a task
description with cycles. All build systems discussed in this paper are correct
only under the assumption that the given task description is acyclic. This
includes the \hs{busy} build system introduced earlier: it will loop
indefinitely given a cyclic \hs{tasks}. Some build systems provide a limited
support for cyclic tasks, see~\S\ref{sec-iterative-compute}.
